---
title: "95% Confidence Intervals on a point estimate"
author: "Chrissy h Roberts"
---

## Find the 95% CI on a point estimate

This simple approach is based on the formula

$\LARGE p ± Z * \sqrt\frac{p(1-p)}{n}$

Where

$\LARGE p$ = point estimate

$\LARGE Z$ = Zcrit value for 95% confidence level (i.e. 1.96 for a 95% confidence interval)

$\LARGE n$ = sample size

## Libraries

```{r}
library(tidyverse)
library(knitr)
```

Define Some data

```{r}
df = tibble(
            month = 1:10,
            prevalence = c(0.72,0.62,0.44,0.22,0.17,0.12,0.13,0.09,0.04,0.02)
            )

kable(df)
```

## Define a function to calculate upper and lower confidence interval

```{r}
point.estimate.CI <- function(p,z=1.96,n){z * sqrt((p*(1-p))/n)}
```

### Capture the upper and lower limit for a given value of n

```{r,warning=F,message=F}

df<-df %>% 
  mutate(
        upper10 = prevalence + point.estimate.CI(prevalence,n = 10),
        lower10 = prevalence - point.estimate.CI(prevalence,n = 10),
        upper50 = prevalence + point.estimate.CI(prevalence,n = 50),
        lower50 = prevalence - point.estimate.CI(prevalence,n = 50),
        upper1000 = prevalence + point.estimate.CI(prevalence,n = 1000),
        lower1000 = prevalence - point.estimate.CI(prevalence,n = 1000)
        ) 
```

### Draw the confindence intervals

This chart shows the point estimates (black dots) as well as the 95% confidence intervals obtained when n was 10 (green ribbon), 50 (red ribbon) or 1000 (blue ribbon)

```{r}
ggplot(df,aes(month,prevalence))+
  geom_ribbon(aes(x = month,y=prevalence,ymin=lower10,ymax=upper10),alpha=0.4,fill="green")+
  geom_ribbon(aes(x = month,y=prevalence,ymin=lower50,ymax=upper50),alpha=0.6,fill="red")+
  geom_ribbon(aes(x = month,y=prevalence,ymin=lower1000,ymax=upper1000),alpha=0.6,fill="blue")+
  geom_point()+
  geom_line()

```
