---
title: "Extract Model from geom_smooth model"
author: "Chrissy h Roberts"
---

Example code to extract x and y values from a geom_smooth.

The practical example used her is to take discrete values of a value n at monthly intervals across a year. The geom smooth is used to draw a smoothed line across the chart. Then the values of n are extracted from the model before being floored so that we can estimate a new value for each month.

This would be helpful where you have a situation in which there's a month missing in the middle, or where there's an outlier or two. You could achieve the same thing using a regression, but this is a useful approach that ultimately leans back on to polynomial regressions.

```{r,warning=F,message=F}
library(tidyverse)
library(knitr)
```

Make a dummy data set with 11 observations and increasing number

```{r,message=F}
df<-tibble(
           month=1:11,
           n = c(100,140,200,260,360,470,560,630,770,990,1100)
           )
kable(df)
```

Make a ggplot object with a geom_smooth()

```{r,warning=F,message=F}
p <- ggplot(df, aes(month, n))+geom_smooth(method = "lm") +geom_point()
p
```

Use ggplot_build to find the code behind the curve - the first element of data in the result is the geom_smooth curve

```{r,message=F,warning=F}
pp<- ggplot_build(p)$data[[1]] %>% 
  rename (
    month = x,
    n = y
  )
kable(pp)
```

Plot the data from what we just extracted, to prove the principle.

```{r,message=F}


ggplot() + 
  geom_smooth(aes(month, n), df,color="blue",lwd=2)+
  geom_line(aes(month, n), pp,color="red",lwd=1)

```

To find a start of month amount, find the floor of each integer on x axis

```{r}
pp$month<-floor(pp$month)
```

Remove replicates and select the values of x and y

```{r,message=F}
pp <- pp %>% mutate(dup = duplicated(month)) %>% 
             filter(dup==FALSE) %>% 
             select(month,n)

df<-bind_cols(df,pp[,2])
kable(df)
```
