---
title: "ODK GPS to KML"
author: "Chrissy h Roberts"
---

## Background

Keyhole Markup Language (KML) is an XML notation for expressing geographic annotation and visualization within two-dimensional maps and three-dimensional Earth browsers. KML was developed for use with Google Earth, which was originally named Keyhole Earth Viewer. Some users of ODK will like to view their GPS data as a Google Earth layer.

This script converts basic GPS data to a KML file

## Libraries

```{r}
library(tidyverse)
library(sp)
library(rgdal)
```

## Dummy data

This data will resemble the kind of lat/long data that comes off ODK. I've supplemented it with some annotations and a 'status' which will colour a flag or something.

```{r}
df<- tibble(
          lon=c(-123.736597,6.263428,147.263399,83.263430,169.263481,7.736540,0.0)  
          lat=c(21.04847,-66.04859,42.04850,66.04854,44.04862,50.04865,52.00000)
          status=c(1,1,1,1,2,2,2)
          names=("Mid_Pacific_Off_Mexico",
                 "North_Of_Antarctica",
                 "East_Of_Hokkaido",
                 "North_Central_Russia",
                 "Suiko_Seamount",
                 "Bacharach_Germany",
                 "Barkaway_UK")
            )
```

## Set up the geometry of the data set

```{r}
#set the coordinate columns
coordinates(df) <- ~lon+lat
#set the geometry of the earth that is used in the map. This is set for standard GPS method
proj4string(df) <- CRS("+init=epsg:4326 +proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")

```

## Create the KML

```{r}
##export basic KML
kml(df)

#with labels
kml(df,labels = names)

#add a shape that can be coloured by another variable like status
shape = "http://maps.google.com/mapfiles/kml/pal2/icon18.png"
#list of icons here https://sites.google.com/site/gmapsdevelopment/

# color only:
kml(df, shape = shape, colour = status, labels = names)

# change size
kml(df, shape = shape, size = 5, colour = status, labels = names)

# two aesthetics elements are effective in emphasizing hot-spots:
kml(df, shape = shape, altitude = 100, extrude = TRUE, 
    colour = status, labels = names, kmz = TRUE)
```
