---
title: "Hazard Ratio Chart from Summary Data"
author: "Chrissy h Roberts"
---

## Background

Often a user just wants to draw a quick chart of study results using only summary data. For instance, you may be looking at a table of hazard ratios and their confidence intervals and would like to visualise the results instead of working with the table.

This short script can take a vector of estimates, along with 95% CIs and will draw a chart.

## Libraries

```{r}
library(tidyverse)
```

## Dummy Data

Start by building a data frame using *tibble.* You should provide some labels (here IMD), along with an estimate (ABSOLUTE.CHANGE) and both lower (LCI) and upper (UCL) confidence limits

```{r}

df <- tibble( 
            IMD = c("IMD1","IMD2","IMD3","IMD4","IMD5"),
            ABSOLUTE.CHANGE  = c(-3.1,-6.6,-1.6,-2.0,-0.9),
            LCI = c(-0.4,-4.4,0.49,-1.06,-0.2),
            UCI = c(-5.7,-8.8,-3.62,-2.9,-1.5)
)

df
```


## Chart

There's nothing clever about this chart. It uses geom_point to draw the estimates, then uses geom_errorbar to add the upper and lower confidence limits.

```{r}
ggplot(df,aes(x=IMD,y=ABSOLUTE.CHANGE,colour=IMD))+
       geom_errorbar( mapping=aes(x=IMD, y=ABSOLUTE.CHANGE, ymin=LCI, ymax=UCI), width=0.1, linewidth=1)+
       geom_point(size=5)+
       geom_hline(yintercept = 0,lty=2)  +
       ylim (-10,10)
```

If you have some point estimates and know the sample size, but don't have confidence intervals, then you can calculate the confidence intervals using \[this method\](../examples/95CI_point_estimate.qmd)
