---
title: "Genetic Power Calculator"
author: "Chrissy h Roberts"
---

## Background

These power calculations are useful for genetic case control studies They use the gap package and pbsize2 command

citation Jing Hua Zhao (2013). gap: Genetic Analysis Package. R package version 1.1-10.

## Libraries

```{r}
library(gap)
library(knitr)
```

## Usage

### Define study characteristics

```{r}

prevalence=0.1
alpha=0.05
num.cases=2000
num.controls=2000
num.total=num.cases+num.controls
prop.cases=num.cases/num.total

```

### Define ranges of minor allele frequency and effect sizes

```{r}
maf_values<-seq(0.01,0.49,by=0.01)
risk_values<-seq(1,1.6,by=0.01)
```

### Calculate power at all possible combinations

```{r}
x<-matrix(nrow=length(risk_values),ncol=length(maf_values))

for (i in 1:length(risk_values)){risk = risk_values[i];for (j in 1:length(maf_values)){maf = maf_values[j];try(x[i,j]<-pbsize2(N=num.total,fc=prop.cases,alpha=alpha,gamma=risk,p=maf,kp=prevalence,model="additive"))}}

```

### Define a matrix which contains the results

```{r}

xx<-as.data.frame(x,row.names=as.character(risk_values));names(xx)<-as.character(maf_values)

kable(xx)

```

### Define a palette for the results

```{r}
#set the colours used for the plot
wut.colors<-rev(c("red","magenta","pink","orange","yellow","green","dark green", "blue","purple","dark blue"))
```

```{r}
#print plot of power

image(risk_values,maf_values,as.matrix(xx),zlim=c(0,1),col= wut.colors,breaks=seq(0,1,by=0.1),useRaster=T,main=paste("Power @ Prevalence :",prevalence,"| Cases :",num.cases,"| Controls :",num.controls,"| alpha :",alpha),xlab="Relative Risk",ylab="Minor allele frequency",cex.main=0.8);legend(x=1,y=0.4,cex=0.5,legend=c("0.00-0.09","0.10-0.19","0.20-0.29","0.30-0.39","0.40-0.49","0.50-0.59","0.60-0.69","0.70-0.79","0.80-0.89","0.90-1.00"),fill=wut.colors,title="Power")

```
