---
title: "ruODK Setup and basic functions"
---

## Background

ruODK is a great package for interacting with ODK Central via R. Full details of how to use it are [here](https://github.com/ropensci/ruODK).

To connect your ODK Central account to R, you'll need to set up your passwords in the *r.environ* file so they're not exposed in scripts. Your passwords will be stored in plain text on your *r.environ* so be cautious about which machines you work with this way. There's also options to add you password manually each time, but we'll assume you're on a secure machine when you do this.

## Add your credentials to *r.environ*

Run this command to open the *r.environ* file

> **usethis::edit_r\_environ(scope = "user")**

Inside renviron file, be careful not to delete anything, but add these lines (substituting your ODK central URL and email address in the appropriate places.

> **ODKC_URL="https://central.xxx.com"**
>
> **ODKC_UN="mustafa.orbach\@xxx.com"**

On a new line, add the password you use to log in to the system. Substitute *xxx* for your password

> **ODKC_PW="xxx"**

On a new line, add your password for decryption (if you use project level encryption). Substitute *xxx* for your password

> **ODKC_PP="xxx"**

Finally, save the *r.environ* file and restart R to load it in to R.

## Using ruODK to download a data set

### Libraries

```{r}
library(ruODK)
library(tidyverse)
library(knitr)

```

### Connect to server

My servers are at central.lshtm.ac.uk. You need to change this to yours. **Sys.getenv()** pulls the passwords from the *r.environ* file in a way that is more secure than just adding to the script.

```{r}
#connect to server
ruODK::ru_setup(
  url = "https://central.lshtm.ac.uk",
  un = Sys.getenv("ODKC_UN"),
  pw = Sys.getenv("ODKC_PW"),
  tz = "Europe/London",
  verbose = TRUE
)

```

### Show a list of projects

You'll want to know the pid number, given here as `id`

```{r}
kable(ruODK::project_list()[1:2,-9])
```

### Show a list of forms that are included in the project

```{r}
kable(ruODK::form_list(pid = 37))

```

### Show details about a specific form

```{r, warning=FALSE, r,message=F}
kable(ruODK::form_detail(pid = 37,fid = "geopoint_map"))
```

### Show a list of submissions to a specific form

```{r, warning=FALSE, message=F}
kable(ruODK::submission_list(pid = 37,fid = "geopoint_map"))
```

### Pull submissions

ruODK is very powerful and can get quite complicated. The most basic thing you need is a dataframe with your submissions. Here's how to get one.

```{r, warning=FALSE,message=FALSE,error=FALSE }
df<-ruODK::odata_submission_get(pid = "37",fid="geopoint_map")
```

### Show data

```{r}
kable(df[,1:4])
```

