---
title: "UK Postcode Maps"
author: "Chrissy h Roberts"
---

## Background

This is a basic tutorial on how to load shapefiles and geopoints to a leaflet map.

There is a lot more detailed information here \[https://rstudio.github.io/leaflet/\](https://rstudio.github.io/leaflet/)

The example is hopefully useful as it uses UK postcode and district shapefiles, which are commonly useful for a variety of purposes.

UK Postcode data can be found here

<https://www.freemaptools.com/map-tools.htm>

UK district shapefiles are available here

<https://www.ordnancesurvey.co.uk/opendatadownload/products.html>

## Libraries

```{r}
library(tidyverse)
library(sf)
library(knitr)
library(leaflet)
```

## Data

### Create a folder to house data

```{r}
if(!dir.exists("data/ukpostcodes"))(dir.create("data/ukpostcodes"))
system("rm -rf data/ukpostcodes/*")
```

## Download a list of UK postcodes and gps locations

```{r}
download.file(url = "https://data.freemaptools.com/download/full-uk-postcodes/ukpostcodes.zip",destfile =paste("data/ukpostcodes/ukpostcodes.zip"))

system ("unzip data/ukpostcodes/ukpostcodes.zip -d data/ukpostcodes/")
  
```

## Download a list of UK district shapefiles

```{r}
# set a timeout proportional to the size of the dataset and the speed of the connection
options(timeout=300)

download.file(url = "https://api.os.uk/downloads/v1/products/BoundaryLine/downloads?area=GB&format=ESRI%C2%AE+Shapefile&redirect",destfile =paste("data/ukpostcodes/bdline_essh_gb.zip"))

system ("unzip data/ukpostcodes/bdline_essh_gb.zip -d data/ukpostcodes/")
  
```

## Read Postcode data

```{r}
ukpostcodes <- read_csv("data/ukpostcodes/ukpostcodes.csv")
```

## Make a small sample of all postcodes

```{r}
sample<-ukpostcodes[sample(1:nrow(ukpostcodes),size = 100),] %>% 
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )
head(sample)

```

## Read in shapefiles to a map

```{r}
map = read_sf("data/ukpostcodes/Data/GB/county_region.dbf")

```

```{r}
mapproj <- st_transform(map, st_crs("+proj=longlat +init=epsg:4326 +ellps=WGS84 +datum=WGS84 +no_defs"))
```

```{r,warning=F}
leaflet(mapproj) %>% 
    addTiles() %>% 
  addMeasure(primaryLengthUnit="kilometers", secondaryLengthUnit="meters")  %>%
  addScaleBar(options = c(imperial = FALSE)) %>%  
  addPolygons(color = "green",label = mapproj$NAME) %>% 
  addCircleMarkers(lng = sample$longitude,lat = sample$latitude,label = sample$postcode,radius = 0.3)

```

### remove data

```{r}
system("rm -rf data/ukpostcodes/")
```
