---
title: "Univariate regressions, reported as a table"
author: "Chrissy h Roberts"
---

## Univariate Regressions - reported as a table

The **tbl_uvregression()** function produces a table of univariate regression models. The function is a wrapper for **tbl_regression()**, and as a result, accepts nearly identical function arguments. The function's results can be modified in similar ways to tbl_regression().

This example uses the built-in *trial* data set

```{r,message=F}
library (gtsummary)
library(tidyverse)
library(knitr)
```

Show the data set

```{r}
kable(head(trial,50))
```

Perform univariate analyses

```{r,message=F}
trial %>%

  select(response, age, grade) %>%

  tbl_uvregression(
    method = glm,
    y = response,
    method.args = list(family = binomial),
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 2)
  ) %>%
  add_global_p() %>%  # add global p-value 
  add_nevent() %>%    # add number of events of the outcome
  add_q() %>%         # adjusts global p-values for multiple testing
  bold_p() %>%        # bold p-values under a given threshold (default 0.05)
  bold_p(t = 0.10, q = TRUE) %>% # now bold q-values under the threshold of 0.10
  bold_labels()
```
