---
title: "Normalise utf8 accented text"
author: "Chrissy h Roberts"
---

## Background

Accented text can cause problems for R.

This method shows how to convert all accents from mixed encodings to UTF-8 text .

Problems like this often emerge with data shared between PC/Mac/Linux and with data exported from Excel. The simple solution is to use the ***Encoding()*** function from the ***utf8*** package.

### Libraries

```{r}
library(utf8)
```

### Dummy data

Here we define a vector x, which has accents. To simulate the problem, we set encoding to be mixed between UTF-8 and bytes, but the second entry is actually encoded in Latin *byte* format with the leading byte *0xE7 rather than in UTF-8.*

```{r}

x <- c("fa\u00E7ile", "fa\xE7ile", "fa\xC3\xA7ile")
Encoding(x) <- c("UTF-8", "UTF-8", "bytes")
```

If we try to convert all entries in the vector to UTF-8, it fails

```{r}
try(as_utf8(x))
```

The simple fix is to change the encoding to match the real data. Here entry two is switched to the correct encoding and we are then able to re-encode it

```{r}
Encoding(x[2]) <- "latin1"
as_utf8(x)


```
